PoPoolation Pipeline

### 2022-01-14 poolseq.Rmd
formatting changes through line 39

The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289). Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/). To download this dataset, go to: https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=run_browser , then type in the accession number for the identified files you want to download (e.g., the accession no. for the file Bodega population file is SRR10760095, identified under the "Run" heading). Then click on the Data access tab and there will be a downloadable link. Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER

***Strengths and weaknesses of Pool-seq***
Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information. Therefore, analyses rely on allele frequency estimates rather than hard “genotype calls”. PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome. The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution). For example, to identify small changes in allele frequencies, you would need to genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding. However, individual genotype information is lost, as well as information on dominance and
effect sizes. For example, individual genotypes cannot be linked to individual phenotypes. Additionally, estimations of linkage disequilibrium are limited to a single sequencing read since that is the known limit of haplotype information for a single individual. Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited. Unlike sequencing of individuals, this cannot be solved by analyzing multiple reads from the same region of a single chromosome. One way to ameliorate this issue is to sequencing replicates of pools. In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods. Care must be taken to ensure the the experimental design maximizes the ratio sequencing coverage to the number of individuals pooled.

***Notes on Experimental Design for Downstream Analyses:***
It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample. Robert Kofler and team also recommend that the pool size should be larger than the goal coverage for sequencing, because this minimizes re-sampling the same allele from a single individual several times. Because PoPoolation is designed for comparing differences in allele frequencies among treatments/population, accuracy will depend on the pool size and sequencing coverage. For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage. The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient. If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
The sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig. 1a from Schlötterer et al., 2014).

@@

***Alternative Pipelines:***
Low Coverage Pool-seq Data: 
https://github.com/petrov-lab/HAFpipe-line
Alternative pipelines have been created for low coverage (<5x) pool-seq data that can provide accurate estimates of allele frequencies. This pipeline is designed for Evolve and Re-sequence studies where the founder haplotypes are known.

Customizable Filtering and Assessing Quality of Pool-Seq Data:
https://github.com/ToBoDev/assessPool
This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data. It determines pool number and prepares proper data structure for analysis, creates a customizable run script for Popoolation2 for all pairwise comparisons, runs Popoolation2 and poolfstat, imports Popoolation2 and poolfstat output, and finally, generates population genetic statistics and plots for data visualization.

>>>>

The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289). Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/). To download this dataset, go to: https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=run_browser , then type in the accession number for the identified files you want to download (e.g., the accession no. for the file Bodega population file is SRR10760095, identified under the "Run" heading). Then click on the "Data access" tab, and there will be a downloadable link. Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER

***Strengths and weaknesses of Pool-seq***\
Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information. Therefore, analyses rely on allele frequency estimates rather than hard “genotype calls”. PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome. The pooling of many individuals can be a cost-effective tool to characterize genetic variation present at the population level (e.g., genome wide association studies and experimental evolution). For example, to identify small changes in allele frequencies, you would need to genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding. However, individual genotype information is lost, as well as information on dominance and
effect sizes. For example, individual genotypes cannot be linked to individual phenotypes. Additionally, estimations of linkage disequilibrium are limited to a single sequencing read since that is the known limit of haplotype information for a single individual. Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited. Unlike sequencing of individuals, this cannot be solved by analyzing multiple reads from the same region of a single chromosome. One way to ameliorate this issue is to sequence replicates of pools. In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods. Care must be taken to ensure the the experimental design maximizes the ratio sequencing coverage to the number of individuals pooled.

***Notes on Experimental Design for Downstream Analyses:***\
It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample. Robert Kofler and team also recommend that the pool size should be larger than the goal coverage for sequencing, because this minimizes re-sampling the same allele from a single individual several times. Because PoPoolation is designed for comparing differences in allele frequencies among treatments/population, accuracy will depend on the pool size and sequencing coverage. For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage. The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences:\"To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient. If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."\
It is critical to note, the sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig. 1a from Schlötterer et al., 2014).

@@

***Alternative Pipelines***\
Low Coverage Pool-seq Data:\ 
https://github.com/petrov-lab/HAFpipe-line \
Alternative pipelines have been created for low coverage (<5x) pool-seq data that can provide accurate estimates of allele frequencies. This pipeline is designed for Evolve and Re-sequence studies where the founder haplotypes are known.

Customizable Filtering and Assessing Quality of Pool-Seq Data:\
https://github.com/ToBoDev/assessPool \
This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data. It determines pool number and prepares proper data structure for analysis, creates a customizable run script for Popoolation2 for all pairwise comparisons, runs Popoolation2 and poolfstat, imports Popoolation2 and poolfstat output, and finally, generates population genetic statistics and plots for data visualization.
